---
title: "Bar, dot, and upset plots"
author: "David A. Morales-Vicente"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE,message=F,warning=F}
knitr::opts_chunk$set(echo = TRUE)
```

## Generation of dendrogram of clusters

To add the dendrogram of cell-clusters to the y-axis of bar, and dot plots it is necessary first build one based on the expression data. We used the median expression data of all cells from the same cell-cluster as the input to calculate the distance and relationship between clusters in the dendrogram.

```{r dendrogram, eval=T,echo=T,include=T,message=F,warning=F}

library(Seurat)
library(tidyverse)
library(magrittr)
library(ggdendro)

# Loading sparse expression matrix from the RDS object of the current analysis

sm_integrated = readRDS("sm_all_genes.rds")
sm_sp_matrix = sm_integrated[["SCT"]]@counts

# Making table of embedding and cell types from current analysis

cell_id  = sm_integrated$orig.ident %>%
              enframe(name = "cell_id",value = "individual") %>%
              mutate(group = str_remove(individual,".$"))

cell_id = sm_integrated$predicted.id %>%
              enframe(name = "cell_id",value = "ref_cluster") %>%
              inner_join(cell_id)

cell_id = sm_integrated$seurat_clusters %>%
              enframe(name = "cell_id",value = "seurat_cluster") %>%
              inner_join(cell_id)

umap_emeddings = sm_integrated@reductions$ref.umap@cell.embeddings %>%
                  as_tibble(rownames = "cell_id")

cell_id %<>% inner_join(umap_emeddings) %>% arrange(cell_id)

# Making list of cells by cell-cluster annotation

cell_groups = split(cell_id$cell_id,cell_id$ref_cluster)

# Calculating the median of the expression of the cells by cell-clusters

median_exp_cluster = t(sapply(cell_groups,
                              function(i) apply(sm_sp_matrix[,i ,drop=F],1,median)))

#Working only in genes identifies as markers

sm_markers_clusters = read_tsv("sm_markers_ref_clusters.tsv")

markers = sm_markers_clusters %>%  ungroup() %>% select(gene) %>% unique() %>% .$gene

median_exp_cluster = median_exp_cluster[,markers]

#Making dendrogram

dd = dist(median_exp_cluster, method = "minkowski")
hc = hclust(dd, method = "ward.D")

plot_levels = hc$labels[hc$order]

# Plotting dendrogram

ddg = ggdendrogram(hc,rotate = F) + ggtitle("Median Cluster method ward.D")
print(ddg)
```

## Matching the cell names between datasets 

To calculated the percentage of cells that remapped to the same cell-cluster it was necessary to match the cell names from the current analys to the cell names from the RDS object from Wendt. et al 2020. 


```{r renaming, echo=F,include=T,message=F,warning=F}

# Loading Wendt et al. RDS object

sm_wendt = readRDS("/work/users/david_morales/schisto/scRNA/collins/publicated/GSE146736_adult_scseq_seurat.rds")
sm_wendt$cell_types = sm_wendt@active.ident

wendt_cell_id = sm_wendt$cell_types %>% 
                  enframe(name = "cell_id",value = "wendt_cluster")  %>% 
                  arrange(cell_id)

# Correcting the cell names from the current analysis to have the same cell names from Wendt et al. data

wendt_barcodes = wendt_cell_id %>% mutate(barcode = str_match(cell_id,"(.*)?_")[,2],
                                          number =  str_match(cell_id,"_(.*)$")[,2])  %>%
                      group_by(barcode) %>%
                      summarise(wendt_n = paste(number,collapse = ","))

cell_barcodes = cell_id %>% mutate(barcode = str_match(cell_id,"(.*)?_")[,2],
                                   number =  str_match(cell_id,"_(.*)$")[,2]) %>%
                  group_by(barcode) %>%
                  summarise(number = paste(number,collapse = ","))


wendt_to_n = cell_barcodes %>% full_join(wendt_barcodes) %>%
                filter(!is.na(wendt_n) & !is.na(number)) %>%
                group_by(number) %>%
                count(wendt_n) %>% 
                filter(!( str_detect(number,",") | str_detect(wendt_n,","))) %>%
                arrange(number,desc(n)) %>%
                mutate(rank = 1:length(n)) %>%
                filter(rank == 1) %>% select(number,wendt_n) %>% ungroup()

# Changing the name to the Wendt et al. format

cell_id %<>% mutate(barcode = str_match(cell_id,"(.*)?_")[,2],
                              number =  str_match(cell_id,"_(.*)$")[,2])  %>%
              inner_join(wendt_to_n) %>%
              mutate(old_cell_id = cell_id,
                     cell_id = paste(barcode,wendt_n,sep="_")) %>%
  left_join(wendt_cell_id)

```

```{r pressure, echo=T,eval=F,include=FALSE,message=F,warning=F}
# Loading Wendt et al. RDS object

sm_wendt = readRDS("/path/to/GSE146736_adult_scseq_seurat.rds")
sm_wendt$cell_types = sm_wendt@active.ident

wendt_cell_id = sm_wendt$cell_types %>% 
                  enframe(name = "cell_id",value = "wendt_cluster")  %>% 
                  arrange(cell_id)

# Correcting the cell names from the current analysis to have the same cell names from Wendt et al. data

wendt_barcodes = wendt_cell_id %>% mutate(barcode = str_match(cell_id,"(.*)?_")[,2],
                                          number =  str_match(cell_id,"_(.*)$")[,2])  %>%
                      group_by(barcode) %>%
                      summarise(wendt_n = paste(number,collapse = ","))

cell_barcodes = cell_id %>% mutate(barcode = str_match(cell_id,"(.*)?_")[,2],
                                   number =  str_match(cell_id,"_(.*)$")[,2]) %>%
                  group_by(barcode) %>%
                  summarise(number = paste(number,collapse = ","))


wendt_to_n = cell_barcodes %>% full_join(wendt_barcodes) %>%
                filter(!is.na(wendt_n) & !is.na(number)) %>%
                group_by(number) %>%
                count(wendt_n) %>% 
                filter(!( str_detect(number,",") | str_detect(wendt_n,","))) %>%
                arrange(number,desc(n)) %>%
                mutate(rank = 1:length(n)) %>%
                filter(rank == 1) %>% select(number,wendt_n) %>% ungroup()

# Changing the name to the Wendt et al. format

cell_id %<>% mutate(barcode = str_match(cell_id,"(.*)?_")[,2],
                              number =  str_match(cell_id,"_(.*)$")[,2])  %>%
              inner_join(wendt_to_n) %>%
              mutate(old_cell_id = cell_id,
                     cell_id = paste(barcode,wendt_n,sep="_")) %>%
  left_join(wendt_cell_id)

```

## Bar plot
### Figure 2

Using the data from Wendt et al. 2020 <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE146736> the percentage of cells that remapped to the same cell-cluster was calculated. 

```{r,fig.height = 14, fig.width = 8, fig.align = "center",message=F,warning=F}

library(ggh4x)

# Counting the number of cells for each cell-cluster 

n_ref_cell_id = wendt_cell_id %>% count(wendt_cluster) %>% 
                  rename(Wendt = n,
                         ref_cluster = wendt_cluster)

n_cell_id = cell_id %>% count(ref_cluster) %>% rename(remapping = n)


# Counting the number of cells that remapped to the same cell-cluster

p_right_remap = cell_id %>% 
                  filter(wendt_cluster == ref_cluster) %>%
                  count(ref_cluster) %>% inner_join(n_ref_cell_id) %>%
                  mutate(p = n/Wendt*100,
                         type = "remapped_n") %>%
                  rename(remapping = n)

# And all the number of cell that belong to a cell-cluster

n_ref_cell_id = n_ref_cell_id %>% inner_join(n_cell_id) %>% 
                  mutate( p = remapping/Wendt*100,
                          Wendt = Wendt/Wendt*100) %>% 
                  mutate(type = "final_n")

p_cell_to_clusters = n_ref_cell_id %>% bind_rows(p_right_remap)


# Making the plot

p_cell_to_clusters_g = n_ref_cell_id %>%
                          ggplot(aes(p,ref_cluster)) +
                          geom_col(fill = "black",colour = "black",
                                   show.legend = F) +
                          theme_classic() +
                          xlab("Percentage of cells related to Wendt et al") +
                          scale_x_continuous(breaks = c(25,50,75,100,125,
                                                        150,175,200,250,300)) +
                          scale_y_dendrogram(hclust = hc,drop=FALSE) +
                          theme(axis.title.y = element_blank(),
                                axis.title = element_text(size =14),
                                axis.text.x = element_text(size =14),
                                axis.text.y = element_text(size = 14))


p_cell_to_clusters_g = p_cell_to_clusters_g + 
                          geom_col(data= p_right_remap,aes(p,ref_cluster),
                                   fill = "white",colour = "black") + 
                          geom_vline(xintercept = 100,
                                     linetype = "dashed",colour = "red") +
                          geom_text(aes(label = remapping),nudge_x = 15,size = 4)+
                          geom_text(data = p_right_remap,aes(label = remapping),
                                    position = position_stack(vjust = .5),size = 4) 


print(p_cell_to_clusters_g)

```

## Finding the stringent markers

Making the tables containing expression data to use them later for the plots.

```{r markers, message=F, warning=F}

#Loading the table of DE genes by cell-cluster

sm_markers_clusters = read_tsv("sm_markers_ref_clusters.tsv")

# Selecting only the relevant fields of the table and subsetting the lncRNA genes
sm_markers_sum = sm_markers_clusters %>%
                    select(cluster,gene,avg_log2FC,p_val_adj,gene_type) %>%
                    rename(ref_cluster = cluster)

lncRNA_markers = sm_markers_clusters %>% 
                      filter(gene_type == "lncRNA") %>%
                      ungroup() %>% select(gene) %>% unique() %>% .$gene


## Changing the cell ids names in the SCT sparse matrix to the new cell id

sm_sp_matrix = sm_integrated[["SCT"]]@counts
sm_sp_matrix_t = Matrix::t(sm_sp_matrix)

cell_id = rownames(sm_sp_matrix_t) %>% 
           enframe(name = NULL,value = "old_cell_id") %>% inner_join(cell_id)

rownames(sm_sp_matrix_t) = cell_id$cell_id

# Making a table with the SCT gene count information

gene_counts = sm_sp_matrix_t[,colnames(sm_sp_matrix_t) %in% lncRNA_markers] %>%
                    as_tibble(rownames = "cell_id") %>%
                    pivot_longer(-cell_id,names_to = "gene",values_to = "counts") %>%
                    inner_join(cell_id) %>%
                    arrange(ref_cluster)

# Summarizing data for further selection

lncRNA_counts_sum = gene_counts %>%
                      group_by(ref_cluster,gene) %>%
                      summarise(n = sum(counts>0),
                                t = length(counts),
                                counts = median(counts),
                                p = n/t*100)

lncRNA_counts_all = gene_counts %>%
                      group_by(gene) %>%
                      summarise(n = sum(counts>0),
                                t = length(counts),
                                counts_all = median(counts),
                                p_all = n/t*100) %>%
                      select(gene,counts_all,p_all)

lncRNA_counts_sum %<>% inner_join(lncRNA_counts_all) %>%
                    mutate(p_ratio = p/p_all) %>%
                    select(ref_cluster,gene,counts,counts_all,contains("p"))
```

## Dot plot
### Figure 3

Dotplot of the lncRNA markers showing the percentage of cells expressing a lncRNA and the median expression of that lncRNAs in a cell-cluster.

```{r,fig.height = 14, fig.width = 17, fig.align = "center", message=F, warning= F}

# Making list of top cell-clusters lncRNA markers, those with a median SCT count greater than 1 in a cell-cluster compare to the median of all the other cells

lcnRNA_top_markers = lncRNA_counts_sum %>%
                      inner_join(sm_markers_sum) %>%
                      filter(counts >= 1)  %>%
                      filter(counts - counts_all >= 1) %>% 
                      mutate(ref_cluster = factor(ref_cluster,levels = plot_levels)) %>%
                      arrange(ref_cluster) %>%
                      .$gene %>% unique()

# Making the dotplot

dot_plot_g = lncRNA_counts_sum %>%
                filter(gene %in% lcnRNA_top_markers) %>%
                mutate(ref_cluster = factor(ref_cluster,levels = plot_levels),
                       gene = factor(gene,levels = lcnRNA_top_markers)) %>%
                
                group_by(gene) %>%
                
                mutate(Scale = scale(counts,center = T)) %>%
                filter(p > 1,ref_cluster != "hes2+") %>% 
                
                ggplot(aes(gene,ref_cluster)) +
                geom_point(aes(size = p,colour = Scale ))  +
                
                scale_size_binned(name = "Percent (%)\nof cells",range = c(0,5)) +
                scale_y_dendrogram(hclust = hc,drop=FALSE) +
                
                cowplot::theme_cowplot() +
                theme(axis.ticks = element_blank()) +
                scale_colour_stepsn(colours = c("grey","blue","purple","yellow","red")) +
                
                theme(axis.text.x = element_text(angle = 90,vjust = 0.5,hjust = 0.5),
                      axis.text = element_text(size = 14),
                      axis.text.y = element_text(size = 13),
                      axis.title = element_blank(),
                      axis.ticks = element_blank(),
                      axis.line = element_blank(),
                      legend.title = element_text(size = 16),
                      legend.text = element_text(size = 16)) 

print(dot_plot_g)

```

## Upset plot
### Figure 4

Upset plot showing the number of shared and unique lncRNA markers in the cell-clusters of the adult worm S. masoni.

```{r,fig.height = 16, fig.width = 8, fig.align = "center",message=F,warning=F}

library(ggupset)

#Picking best lncRNA marker by group and plotting

lncRNA_upset_g = lncRNA_counts_sum %>%
                  inner_join(sm_markers_sum) %>%
  
  # Selection of the lncRNAs with stringent criteria
  
            mutate(ref_cluster = factor(ref_cluster,levels = plot_levels)) %>%
            filter(counts >= 1)  %>%
            filter(counts - counts_all >= 1) %>%
            filter(ref_cluster != "hes2+") %>% # lost cell-cluster after remapping
  
            mutate(score = counts*avg_log2FC) %>%
            arrange(ref_cluster,desc(score)) %>%
            group_by(gene) %>%
            summarise(cluster = list(ref_cluster)) %>%
            ggplot(aes(x = cluster)) +
            geom_bar() +
            scale_x_upset(sets = plot_levels,reverse = F,
                          order_by = "degree",name = "") +
            theme_combmatrix(combmatrix.label.text = element_text(size=16),
                             combmatrix.panel.striped_background.color.one = "grey",
                             combmatrix.panel.striped_background.color.two = "white",
                             combmatrix.label.extra_spacing = 5,
                             text = element_text(size=16)) 
print(lncRNA_upset_g)

```


```{r, echo=F,eval=T}
sessionInfo()
```
