---
title: "UMAP plots"
author: "David A. Morales-Vicente, Ana Carolina Tahira"

output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE, warning=F}
knitr::opts_chunk$set(echo = TRUE)
```

## Generation of a UMAP model from reference data, and tables containing embedding information 

The first step is to generate a UMAP model from Wendt et al. 2020 <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE146736> data and save it, then the embedding for each cell is collected into a table. Finaly a table with the median values for each cell type is generated to add a layer of labels to each cell-cluster.

```{r wend_loading_eval,eval=T,echo=F,message=F}

library(Seurat)
library(tidyverse)
library(magrittr)

sm_wendt = sm_wendt=readRDS("/work/users/david_morales/schisto/scRNA/collins/publicated/GSE146736_adult_scseq_seurat.rds")

sm_wendt$cell_types = sm_wendt@active.ident

# Generating the UMAP model 
sm_wendt = RunUMAP(sm_wendt,dims = 1:80,return.model = TRUE,
                         verbose = T,n.neighbors  = 36, min.dist  =  0.70)

# Saving the embedding, and cell type information for each cell into a table
wendt_cell_id = sm_wendt$cell_types %>% 
                  enframe(name = "cell_id",value = "wendt_cluster")  %>% arrange(cell_id) 

wendt_embedding = sm_wendt@reductions$umap@cell.embeddings %>% 
                    as_tibble(rownames = "cell_id")

wend_umap = wendt_embedding %>% inner_join(wendt_cell_id)

# Making median of UMAP median values for each cell-cluster
wend_umap_med = wend_umap %>%
                group_by(wendt_cluster) %>%
                summarise(umap1 = median(UMAP_1),
                          umap2 = median(UMAP_2))

# Loading RDS from this project

sm_integrated = readRDS("sm_all_genes.rds")

# Making table of embedding and cell types from current analysis

cell_id  = sm_integrated$orig.ident %>%
              enframe(name = "cell_id",value = "individual") %>%
              mutate(group = str_remove(individual,".$"))

cell_id = sm_integrated$predicted.id %>%
              enframe(name = "cell_id",value = "ref_cluster") %>%
              inner_join(cell_id)

cell_id = sm_integrated$seurat_clusters %>%
              enframe(name = "cell_id",value = "seurat_cluster") %>%
              inner_join(cell_id)

umap_emeddings = sm_integrated@reductions$ref.umap@cell.embeddings %>%
                  as_tibble(rownames = "cell_id")

cell_id %<>% inner_join(umap_emeddings) %>% arrange(cell_id)

# Making median of UMAP median values for each cell-cluster

label_position = cell_id %>% group_by(ref_cluster) %>%
                  summarise(umap1 = median(refUMAP_1),
                            umap2= median(refUMAP_2) )


```

```{r wendt_loading_show, eval=F,echo=T, message=F}

library(Seurat)
library(tidyverse)
library(magrittr)

# Loading Wendt et al. RDS object
sm_wendt = readRDS("/path/to/GSE146736_adult_scseq_seurat.rds" )
sm_wendt$cell_types = sm_wendt@active.ident

# Generating the UMAP model 
sm_wendt = RunUMAP(sm_wendt,dims = 1:80,return.model = TRUE,
                         verbose = T,n.neighbors  = 36, min.dist  =  0.70)

# Saving the embedding, and cell type information for each cell into a table
wendt_cell_id = sm_wendt$cell_types %>% 
                  enframe(name = "cell_id",value = "wendt_cluster")  %>% arrange(cell_id) 

wendt_embedding = sm_wendt@reductions$umap@cell.embeddings %>% 
                    as_tibble(rownames = "cell_id")

wend_umap = wendt_embedding %>% inner_join(wendt_cell_id)
wend_umap_med = wend_umap %>%
                group_by(wendt_cluster) %>%
                summarise(umap1 = median(UMAP_1),
                          umap2 = median(UMAP_2))

# Loading RDS from this project

sm_integrated = readRDS("/path/to/sm_all_genes.rds")

# Making table of embedding and cell types from current analysis

cell_id  = sm_integrated$orig.ident %>%
              enframe(name = "cell_id",value = "individual") %>%
              mutate(group = str_remove(individual,".$"))

cell_id = sm_integrated$predicted.id %>%
              enframe(name = "cell_id",value = "ref_cluster") %>%
              inner_join(cell_id)

cell_id = sm_integrated$seurat_clusters %>%
              enframe(name = "cell_id",value = "seurat_cluster") %>%
              inner_join(cell_id)

umap_embeddings = sm_integrated@reductions$ref.umap@cell.embeddings %>%
                  as_tibble(rownames = "cell_id")

cell_id %<>% inner_join(umap_embeddings) %>% arrange(cell_id)

label_position = cell_id %>% group_by(ref_cluster) %>%
                  summarise(umap1 = median(refUMAP_1),
                            umap2= median(refUMAP_2) )

```

## UMAP cell types
### Figure 1

```{r cluster_plot, echo=T, message=F,fig.height = 6, fig.width = 6, fig.align = "center"}
library(ggrepel)
library(pals)

# Generation of a set of colours for dotplots

umap_colours = alphabet2(26)
names(umap_colours) = NULL

umap_colours = rep(umap_colours,times = 3)[1:68]

names(umap_colours) = wend_umap_med$wendt_cluster
umap_colours[umap_colours == "grey70"] = "grey40"

# Making the UMAP plot

umap_plot_g = wendt_embedding %>%
                ggplot(aes(UMAP_1,UMAP_2)) +
                geom_point(size = .5,colour = "grey") +
                geom_point(data = cell_id,aes(refUMAP_1,refUMAP_2,
                                              colour = ref_cluster),
                           show.legend = F,
                           size = .5) +
                scale_colour_manual(values = umap_colours) +
                geom_text_repel(data = label_position,aes(umap1,umap2,
                                                          label = ref_cluster)) +
                theme_classic() +
                ylab("UMAP 2") + xlab("UMAP 1")

print(umap_plot_g)

```

## UMAP expression plots
### Figure 5-8, S1-S3

Using the expression data from the current analysis a function to plot the expression
of each gene or several genes was written, then the function was used to generate the expression plots
showed in the figures.

```{r expression_plots ,echo=T,warning=F,message=F,tidy=T,fig.height = 8, fig.width = 16, fig.align = "center"}
library(cowplot)
library(patchwork)

# Initiating the sparse expression matrix

matrix = sm_integrated[["RNA"]]@counts

# Expression plot function

feature_plot = function(features = NULL,
                          log_transformation = T,point_size = 0.5,
                          plot_cols = c("lightblue","blue","purple","goldenrod2","red"),
                          ...) {
  
  not_found = features[!(features %in% rownames(matrix)) ]
  
  if( length(not_found) != 0 ){
    message(glue::glue("The following genes were not indentified {paste(not_found,collapse=',')}"))
  }
  
  # Susbsetting and transpose sparse expression matrix
  
  features = features[ features %in% rownames(matrix) ]
  matrix = Matrix::t(matrix[features,,drop=F])
  
  if(log_transformation){
    
    plot_m = matrix %>% 
                  as_tibble(rownames = "cell_id") %>%
                  pivot_longer(-cell_id,names_to = "gene",values_to = "counts") %>%
        # Adding the embedding information
                  inner_join(cell_id) %>%
                  group_by(gene) %>%
        # Log transformation  of counts
                  mutate(Expression = log10(counts+1),
                         gene = factor(gene,levels = features)) %>%
                  arrange(gene,Expression)
    
  } else {
    
            plot_m = matrix[,features,drop=F] %>% 
              as_tibble(rownames = "cell_id") %>%
              pivot_longer(-cell_id,names_to = "gene",values_to = "counts") %>%
        # Adding the embedding information
              inner_join(cell_embeddings) %>%
              group_by(gene) %>%

              mutate(Expression = counts,
                     gene = factor(gene,levels = features)) %>%
              arrange(gene,Expression)
    
  }
  
  g = plot_m %>%
    ggplot(aes(refUMAP_1,refUMAP_2)) +
    geom_point(aes(colour = Expression),size = point_size)  +
    facet_wrap(~gene,...) + 
    
    scale_color_gradientn(colours = plot_cols,
                           limits  = c(0,max(plot_m$Expression)),
                                       label = function(x) sprintf("%.2f", x),  
                                       name = "Expression") +
    
    theme_cowplot() +
    theme(axis.title = element_text(face = "bold", size = 18),
          axis.text = element_text(face = "bold", size = 20),
          axis.line = element_line(size = 1.5), 
          legend.title = element_text(angle = 90, hjust = 0.5, 
                                      face = 2, size = 16), 
          legend.text = element_text(face=2, size = 16),
          strip.background = element_blank(),
          strip.text = element_text(size = 20,hjust = 0,face = "bold"),
          legend.position = c(.125,.85)) +
            guides(colour = guide_colourbar(title.position = "left")) +
    xlab("UMAP-1") +
    ylab("UMAP-2")
  
  rm(matrix,plot_m,cell_embeddings)
  gc()
  
  return(g)
  
}

# Printing the plots from figure 5

p5.1 = feature_plot(features = "G39666") # lncRNA target
p5.2 = feature_plot(features = "Smp-073270") # S. mansoni neural marker 7b2

p = p5.1 + p5.2

print(p)

```

```{r, echo=F,eval=T}
sessionInfo()
```
